results

### 2023-02-20 results_work.RMD
added some work on the results, needs more

---
title: "Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults"
author: "Kristian Lian, Daniel Hammarström, Stian Ellefsen, Håvard Hamarsland"
bibliography: resources/ribose.bib
csl: resources/jap.csl
output:
  word_document:
    reference_docx: reference-template-02.docx
editor_options: 
  chunk_output_type: console
---

@@

```{r packages}
library(tidyverse);library(readxl);library(lme4);library(emmeans);library(dplyr);library(cowplot);library(knitr);library(broom);library(rmarkdown);library(tinytex);library(tidyr);library(arsenal);library(magrittr);library(dabestr);library(nlme); library(pbkrtest)
```

## Introduction

Systematic resistance training (RT) increases muscle mass and strength due to repeated mechanical and metabolic stress. Unfortunately, responses to RT vary widely between individuals, with as many as 10-15% showing impaired muscle growth (ALVAREZ18;MANN14;THALACKER13). Such inability to benefit from RT may be explained by genetics, epigenetics, and/or an unfavourable internal physiological milieu (FIGUEIREDO21;HAMMARSTROM20;THALACKER13).<!-- Check these refs, do they correspond to genet epigen and internal mileu? -->
Although RT program modifications such as altered exercise volume or intensity may induce more favorable adaptations in some individuals, such benefits are not uniform between individuals [@hammarstrom.etal_2020].<!-- Add ref for more rt program modifications -->
For example, although changes in training volume generally induces greater benefits of RT, not all individuals benefit from such modifications [@hammarstrom.etal_2020]. Therefore, other means than modification of RT variables seems necessary to optimize individual responses.<!-- The above sentence might be a bit contradictory? -->

Nutritional supplements such as protein and creatine effectively optimize RT adaptations (CERMAK12;LANHERS15;LANHERS17). However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising, as glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis (MUL15;TANAKA18). Furthermore, energy availability is a decisive factor in *de novo* synthesis of ribosomes <!--Add ref that indicate high energy requirement for ribo biog possibly (VONWALDEN19;TANAKA18?  --> which in turn determines muscle growth by increasing the muscles translational capacity (FIGUEIREDO19;VONWALDEN19;TANAKA18;HAMMARSTROM20;STEC16). Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the regulation of ribosome biogenesis.

Transcription of ribosomal RNA (rRNA) by Pol I is considered the limiting step in synthesizing new ribosomes [@Moss_1995]. Multiple signaling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc (KUSNADI15;VONWALDEn19;TANAKA18). mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels (HOPPE09). Importantly, mTOR and its downstream target S6K1 serve as a mediator from insulin signaling in skeletal muscle[@hillier.etal_2000]. mTOR contributes to forming the preinitiation complex (PIC) that marks the initiation of rRNA transcription (VONWALDEN19;FIGUEIREDO19). Parallel to mTORC1 [@Mori_2021; West_2016], c-Myc increases ribosomal biogenesis directly and through transcriptional control of the upstream binding factor (UBF) [@sanijUBFLevelsDetermine2008; Poortinga_2011]. However, interestingly, UBF phosphorylation, required for interaction with the rDNA promoter, is increased by high glucose in a mTORC1 dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells[@Mariappan_2011]. Additionally, independently from UBF, high glucose was shown to lead to chromatin remodeling, which promotes rRNA transcription[@Zhai_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.

The purpose of this investigation was to determine the effect of 
<!-- -->

the relationship between glucose, RT and ribosome biogenesis we recruited 16 healthy young, moderately trained male and female individuals to moderate volume unilateral resistance training for two weeks. They exercised one leg with glucose and protein, and the other leg with placebo and protein ingestion before and after each RT-session. Our main purpose was to investigate whether there are acute effects of glucose- compared to placebo ingestion before and after RT. To assess this purpose, we measured changes in ribosome biogenesis as accumulation of total RNA, expression of the four mature rRNA's (18S, 5.8S, 28S, 5S), 47S pre-rRNA, and levels of proteins associated with transcription initiation (c-Myc, UBF, rpS6?). As secondary outcomes, we measured the development in skeletal muscle performance during the intervention, and skeletal muscle recovery after the final RT session, as temporal changes in maximal isometric and isokinetic torque. We hypothesised that RT would indeed lead to a prominent accumulation of total RNA, increased expression of rRNA and associated proteins. Further, we hypothesised that RT with glucose and protein would induce an enhanced response compared to placebo, in markers of ribosome biogenesis, as well as muscular performance and -recovery.

## Methods

### Materials and methods

All participants were informed about the potential discomforts and risks associated with the study and gave their informed consent before study enrollment. The study was approved by the regional ethical committee (REK, ID nr. 153628), pre-registered at clinicaltrials.gov (Identifier: NCT04545190), and conducted according to the Helsinki declaration.

### Participants

Sixteen healthy male and female participants (20-33yrs, Tab 1) were recruited to the study through social media advertisement and word of mouth, and through completing the selection process (Fig 1). The eligibility criteria were non-smokers and moderately trained (i.e. 2-8 RT sessions per 14 days for the last six months). Exclusion criteria were previous injury leading to impaired strength, inability to perform resistance training, symptoms and a medical record of metabolic disorders including hyperglycemia, i.e. fasting venous plasma glucose ≥6.1 mmol/L and/or 2-hour glucose tolerance ≥7.8 mmol/L, and/or HbA1c >42 mmol/mol. Of the sixteen participants that commenced the intervention, three participants dropped out. One due to sickness and inability to resume, two participants experienced muscular discomfort connected to heavy resistance training (Fig 1). Baseline characteristics (Tab 1) were measured using DXA (Prodigy Advance PA+302047, Lunar, San Francisco, CA, USA) on Day -1, the last day preceding the RT intervention.

### Experimental design

The study was designed as a 12-day double-blinded randomized controlled trial, with a unilateral RT protocol (Fig 2). Participants were randomized to commence with either glucose (GLU, 30g per bolus) or placebo (PLAC, 0g per bolus), and either dominant or non-dominant leg (Fig 1), alternating RT and supplement from one day to the other (Fig 2). Investigators were blinded as the supplement randomization was disclosed only to personnel not involved in any other aspect of the study, and participants were blinded by diluting boluses of glucose (30g glucose) (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (~0.3g Stevia rebaudiana extract) (Steviosa, Soma Nordic AS, Oslo, Norway) in 300ml Fun Light juice/saft (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g GLU vs. 0g GLU). To ensure equal conditions during training sessions, participants exercised and tested at the same time of day, +/- 1hr with the same trainer, and recorded and repeated their daily macro nutrient intake on pairwise consecutive days (Tab 4).

###### **Table 1:** Baseline characteristics of the participants. BMI: body mass index, A = charactaristics per sex, B = characteristics per leg. Values are mean ± standard deviation.

```{r DXA}
# Script: DXA

# Data handling
table.dxa <- read_excel("./data/dxa/ribose_dxa.xlsx") %>%
        select(sex, subject, age, height, weight, leanmasskg, fatmasskg, ffmg) %>%
        #make grams into kg
        mutate(ffkg = ffmg/1000) %>%
        select(sex, age, height, weight, leanmasskg, fatmasskg, ffkg) %>%
  mutate(bmi = (weight) / (height^2) * 10000)

dxa.table <- table.dxa %>%
        pivot_longer(names_to = "variable",
                     values_to = "values", col= age:bmi) %>%
        select(sex, variable, values) %>%
        group_by(sex, variable) %>%
        summarise(m = mean(values, na.rm = TRUE),
                  s= sd(values, na.rm = TRUE)) %>%
        mutate(stat = paste0(round(m, 1), " (", round(s, 1), ")")) %>%
        select(sex, variable, stat) %>%
        pivot_wider(names_from = sex,
                    values_from = stat) %>%
                mutate(variable = if_else(variable %in% c("age"),
                                          "Age (yrs)",
                                          if_else(variable %in% c("fatmasskg"),
                                                  "Fat mass (kg)",
                                                  if_else(variable %in% c("ffkg"),
                                                          "Fat free mass (kg)",
                                                          if_else(variable %in% c("height"),
                                                                  "Height (cm)",
                                                                  if_else(variable %in% c("leanmasskg"),
                                                                          "Lean body mass (kg)",
                                                                          if_else(variable %in% c("weight"),
                                                                                  "Body weight (kg)",
                                                                                  if_else(variable %in% c("bmi"),
                                                                                          "BMI", "")
                                                                                  )))))))

dxa.table %>%
        kable(col.names = c("Variable", "Female (n=7)", "Male (n=9)"))

```{r dxa per leg}
# Script: DXA

# Subject 104, 110 and 111 dropped out and is therefore excluded from the table

#Load dxa data
dxaleg <-  read_excel("./data/dxa/ribose_dxaprleg.xlsx") %>%
        select(subject, sex, leg, leanmassg, fatmassg, totalmasskg) %>%
        #make grams into kg
        mutate(leanmass = leanmassg/1000,
               fatmass = fatmassg/1000)%>%
        select(subject, sex, leg, leanmass, fatmass, totalmasskg)

# Between groups analysis 
dxa.stat <- dxaleg %>%
        select(subject, leg, leanmass, fatmass, totalmasskg)

# Lean mass between legs, within subject 
leanmass.lm <- lmer(leanmass ~ leg + leg:subject + (1|subject), data = dxa.stat)
lm.plot <- plot(leanmass.lm)
lm.sum <- summary(leanmass.lm)
lm.confint <- emmeans(leanmass.lm, specs = ~ "leg")

# Fat mass between legs, within subject
fat.lm <- lmer(fatmass ~ leg + leg:subject +(1|subject), data = dxa.stat)
fat.plot <- plot(fat.lm)
fat.sum <- summary(fat.lm)
fat.confint <- emmeans(fat.lm, specs = ~ "leg")

# Total mass between legs, within subject
total.lm <- lmer(totalmasskg ~leg + leg:subject + (1|subject), data = dxa.stat)
totlm.plot <- plot(total.lm)
totlm.sum <- summary(total.lm)
supsup <- emmeans(total.lm, specs = ~ "leg")

# Table of leanmass, fatmass and total mass in kg pr. leg
dxalegtable <- dxaleg %>%
        select(leg, leanmass, fatmass, totalmasskg) %>%
        pivot_longer(names_to = "variable",
                     values_to = "values", col= leanmass:totalmasskg) %>%
        group_by(leg, variable) %>%
        summarise(m = mean(values, na.rm = TRUE),
                  s= sd(values, na.rm = TRUE)) %>%
        mutate(stat = paste0(round(m, 1), " (", round(s, 1), ")")) %>%
        select(leg, variable, stat) %>%
        pivot_wider(names_from = leg,
                    values_from = stat)

# Markdown table
dxalegtable %>%
        kable(col.names = c("Variable", "Left leg", "Right leg"))

Six RT sessions were conducted with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein), whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose levels). Participants were asked to refrain from resistance- or high-intensity training of the legs from Day -7 and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from other training sources. If participants did conduct exercise training outside of the prescribed protocol, they were asked to ensure equal loading on pairwise consecutive days.

![](figures/flowchart_selection.PNG)

###### **Figure 1:** Flowchart of the selection process. The order in which participants performed the two intervention blocks was determined in a planned randomised fashion.

![](figures/timeline.PNG)

###### **Figure 2.** Overview of the intervention, with 12 days of concomitant dietary intervention and resistance training (RT), preceded by 7 days of testing. T0 = timepoint 0 (testing days prior to intervention), T1/T2 = biopsy prior to resistance training for leg #1 and #2, respectively, T3/T4 post intervention testing of leg #1 and #2, respectively, including post sixth RT session. Test = physical performance test (unilateral 1RM leg press, knee extension, isometric knee extension force, isokinetic knee extension torque) testing on days 4, 5, 8, 9, T3, T4 will not include 1RM leg press and knee extension, Biopsy = microbiopsy, RT#1 = resistance training leg #1, RT#2 = resistance training leg#2. Other tests (D2O, DXA, spit) will be conducted as part of the study as a whole. These are not relevant to this data set, thus will not be described here.

### Dietary intervention

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until two hours after RT, II) from 2.5 hours after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose/placebo only. Glucose/placebo was ingested at three time points in period I: 30 minutes before RT (0830hrs, 30g vs 0g glucose), immediately before RT (0900hrs, 30g vs. 0g glucose), and immediately after RT (~0930hrs, 30g vs. 0g glucose). Protein was ingested 2hrs before RT (0700hrs, 25g) and immediately after RT (~0930hrs, 25g). In the afternoon (1800hrs-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II, registered in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.), to ensure similar premises for resistance training responses between the two legs. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of PRO supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously. The protein supplement was ingested as 25g Whey Protein Isolate boluses (Proteinfabrikken, Stokke, Norway), diluted in 150ml water. During sessions, participants were free to ingest water ad libitum.

### Assessment of muscular strength

Performance tests (Test, Fig 2) were performed prior to (Days -7 and -5, and T0, both legs) and during the intervention (Days 4, 5, 8, and 9), after session 5 and after finalization of the intervention (T3/T4, RT#1 leg; T4/Day 13, RT#2 leg), a total of 10 days (Fig 2). Maximal isometric and isokinetic knee extension torque were measured with a Humac Norm Dynamometer (CSMi, Stoughton, Massechusetts, USA). Individual positions were recorded and standardized from pre-intervention tests (days -7 and -5). Maximal isokinetic torque was measured at speeds of 60- and 240 d/s, 2x3 repetitions each, with the first set of each exercise as a sub-maximal warm-up. The maximal isometric torque was measured at knee-angle 60, for a maximum of 10 seconds and two repetition per test. The highest peak torque values were carried forwards to the final analyses. During days 4, 5, 8 and 9, humac tests were conducted one hour before RT with the leg performing RT the previous day. During days 11 and 12 (T3 and T4), humac tests were performed at four time points: I) 45min before RT, II) 30min after the last RT session, III) 2hrs after the last RT session, and IV) 23hrs after the last RT session. Test I at T4 included testing of both legs, representing 23hrs post RT session test of one leg and pre-session 6 test of the other leg.

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted at pre-intervention testing (days -7 and -5, Fig 2). The participants performed a general warm-up with 10min cycling on an indoor exercise bicycle and a specific warm-up (1x10, 1x6 and 1x3 repetitions, 2min breaks) before each of the tests. Between specific warm-up sets as well as between each subsequent 1RM attempt, the participants were given 3min breaks. All positions were controlled and recorded at pre-intervention testing and repeated for the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, with a knee angle of 45 degrees (start), within a 2.5kg range. Attempts, where participants did not reach 45 degrees during the eccentric phase, were not approved. Maximal knee extension testing followed the same specific warm-up and pause protocol and was defined as maximal load lifted in a controlled fashion, reaching full extension of the knee joint, within a 1.25kg range. Attempts with exaggerated hip movement or beneath full extension were not approved.

### Resistance training protocol

Resistance training consisted of three sets of each unilateral leg presses and knee extensions, with an intensity of 10 repetition maximum (10RM). As a general warm-up, the participants cycled on an indoor exercise bicycle for 5-10min. In addition, before the respective exercises, two 10-repetition warm-up sets were completed at ~50% and ~70% of 10RM. Rest time between working sets was 2min, and progressive loading was used to ensure adequate exercise stimulation throughout the intervention. For safety and standardization purposes, all sessions were monitored by trained personnel. To further ensure equal exercise premises for the two legs, each participant was accompanied by the same trainer in each session. If this was not possible throughout the intervention, each participant had the same trainer on pairwise consecutive days as a minimum. The rate of perceived exertion (RPE, 0–10-point scale) was logged before every session, and the session score was logged 15min after each session. Lastly, training volume (i.e., load and repetitions) was logged for every session.

### Sampling of muscle tissue and blood

Muscle biopsies were sampled from m. vastus lateralis using well-established procedures (HAMMARSTRØM20). Muscle biopsy sampling was performed under local anaesthesia (Xylocaine, 10mg ml-1 with adrenaline 5μg ml−1, AstraZeneca AS, Oslo, Norway) using a 12-14-gauge needle (Universal Plus, Mermaid Medical AS, Stenløse, Denmark), operated with a spring-loaded biopsy gun. After the biopsy sampling, muscle tissue was divided in two aliquots for determination of total RNA/expression of rRNA and two aliquots for protein analyses, snap frozen in isopentane (-80°C) and stored at -80°C until further analyses. Muscle biopsies were sampled at four time points (Fig 2): I/II) Pre-intervention (2hrs before training, time point T1 = pre RT#1 leg, T2 = pre RT#2 leg), and III/IV) before the sixth RT session from RT#1 leg (2hrs before training, T3) and before the sixth RT session from RT#2 leg (2hrs before training, T4). At each time point, two samples were taken from the same incision. To standardize this procedure, all individual participants had biopsies taken at the same time of day, in an overnight fasted state.

To measure blood glucose levels with and without glucose ingestion/training, blood was collected by means of finger draws on days with biopsy sampling. One finger draw was taken at T1 (pre-test, Fig 2) to serve as a baseline. At T3 and T4 (post-test, Fig 2), finger draws were collected at 7 time points: I) prior to protein ingestion (0700hrs) II) 45 minutes after protein ingestion (0745hrs) III) 1.5hrs after protein ingestion (0800hrs, i.e., immediately before GLU/PLAC intake), IIII) 2hrs after protein ingestion (0900hrs, i.e., immediately before training), IV) in the middle of RT (~0915hrs), V) immediately after training (~0930hrs), and VI) 2hrs after completion of training (~1130hrs). Finger draws were analysed with in-house equipment (BIOSEN C-Line, EKF diagnostic GmbH, Barleben). Venous blood samples were collected from the antecubital vein, coinciding with the finger draws except in the middle of the RT session, to analyse endocrine variables.

### Total RNA extraction and quantitative real-time reverse transcription polymerase chain reaction

Two muscle biopsy samples were taken for total RNA extraction per leg per time point, and total RNA was extracted in two duplicates per muscle biopsy. Thus, we had two duplicates per leg per time point, giving a total of eight RNA samples per participant. Total RNA was extracted using TRIzol according to the manufacturer’s protocol. An exogenous RNA control (λ polyA External Standard Kit, Takara Bio Inc., Shiga, Japan) was added at a fixed amount (0.04 ng ml-1 of Trizol reagent) to enable analysis of target gene expression per unit tissue weight (ELLEFSEN2008,2014). Briefly, muscle tissue was homogenized in TRIzol using 0.5mm RNase-free Zirconium beads (~50 ul) (Next Advanced, Averill Park, NY, USA) and a bullet blender (bb). Chloroform (Sigma-Aldrich, Missouri, USA) was used for phase-separation, and an RNA pellet was precipitated with isopropanol (VWR International, Pennsylvania, USA). For assessment of RNA content and -purity 5µl RNA, from the RNA stock, was eluted in TE-buffer (1:2), for assessment via spectrophotometry. All samples had a 260nm to 280nm ratio > 1.9. The RNA stock was stored at -80°C until further analyses. RNA was reverse transcribed using Super Script IV Reverse Transcriptase (Invitrogen), according to the manufacturer’s instructions. Briefly, 1µl 10 mM dNTP mix, 0.5µl anchored oligo-dT, 0.5µl random hexamer primers (Thermo Scientific), a maximum of 9µl template RNA and 2 µl nuclease-free water were mixed, vortexed and briefly spun down. Samples were then heated at 65°C for 5 minutes, followed by at least 1-minute incubation on ice. For the next step, 4µl 5x SSIV Buffer, 1µl 100 mM DTT, 1 µl RNase OUT, and Super Script IV Reverse Transcriptase (Invitrogen) were mixed, vortexed and briefly spun down and added to the samples. Samples were incubated for 10 minutes at 23°C, 10 minutes at 50-55°C and 10 minutes at 80°C. All samples were reverse transcribed and diluted to 1:50 before quantitative real-time polymerase chain reaction (qPCR). Lambda was used as an external reference gene, added in the RNA extraction (2µl per extraction). qPCR reactions were run on a fast-cycling real-time detection system (Applied Biosystems 7500 fast Real-Time PCR Systems, Life Technologies AS), with a total volume of 10 µl, containing 2 µl of cDNA, gene-specific primers (0.5 µM final concentration) and a commercial master mix (2X SYBR Select Master Mix, Applied Biosystems, Life Technologies AS) (Hammarström et al. 2020). qPCR reactions consisted of 40 cycles (3 s 95°C denaturing and 30 s 60°C annealing) (Hammarström et al. 2020). Raw fluorescence data were exported from the platform-specific software and amplification curves were modelled with a best-fit sigmoidal model using the qPCR package (Ritz & Spiess, 2008) written for R (Hammarström et al. 2020; Team 2018). For an overview of primers, see Table 2.

###### **Table 2:** Primer sequence and performance. Average cycle thresholds (Ct) and priming efficiencies were calculated from all qPCR reactions.

![](figures/Primer.sequences.PNG)

### Protein extraction and immunoblotting

Total protein was extracted using the Minute Total Protein Extraction Kit for Muscles (Invent Biotechnology), according to the manufacturer’s protocol, optimised for our lab. Wet muscle was freeze-dried for 24hrs and dissected before extraction. The tissue was homogenized with a plastic rod in 80mg protein extraction powder (Invent Biotechnology) and 100ul iced cell lysis buffer (Denaturing Buffer, Invent Biotechnology), and centrifuged at top speed for 1 min. The supernatant was aliquoted into 2x25ul tubes to run samples in duplicates, and total protein concentrations were determined in a 1:10 dilution (Pierce Detergent Compatible Bradford Assay Reagent, Thermo Fisher Scientific, Oslo, Norway). The protein samples were diluted to 20.25 µg µl−1 total protein with lysis buffer and 4X Laemmli sample buffer (Bio-Rad Laboratories AB, Oslo, Norway) containing 2-mercaptoethanol. Twenty microlitres from each protein sample were pooled as a normalisation factor for gel separation. All protein samples were incubated at 95°C and stored at -20°C until further analysis. The protein samples (20.25µg total protein) were separated at 250V for 50min using 4-20% Tris-Glycine gels (Thermo Fisher Scientific), then transferred to PVDF membranes with wet transfer at 300mA for 3hrs. Both gel electrophoresis and protein transfer were performed at 4°C. Following the wet transfer, membranes were stained with a MemCode reversible total protein stain (MANUFACTURER) and then blocked for 1hr at room temp with a blocking buffer of Tris-buffered saline (TBS; 20 mM Tris, 150 mM NaCl) with 5% non-fat dry milk and 0.1% Tween-20. Primary and secondary antibodies were purchased from Santa Cruz Biotechnology (Texas, USA) UBF:(UBF-9, sc-13125), rpS6:(Ribosomal protein S6 C-8, sc-74459), and Thermo Fisher Scientific () c-Myc:(c-Myc 9E10), goat anti-mouse for c-Myc (goat anti-mouse IgG1 (y1) horseradish peroxidase conjugate), and anti-mouse (anti-mouse IgG1 horseradish peroxidase conjugate). Antibodies were diluted in a 5% non-fat dry milk blocking buffer (TBS, 5% non-fat dry milk, 0.1% Tween-20). Membranes were incubated overnight with primary antibodies, and for 1hr with secondary antibodies. Between blocking and primary antibody staining, membranes were washed 1x5min, between primary and secondary staining, and after secondary staining, membranes were washed 3x5 with TBS-Tween (TBS; 20mM Tris, 150mM NaCl, 0.1% Tween). Following the last TBS-T wash, membranes were incubated 1x5min with enhanced chemiluminescent substrate (ECL, SuperSignal West Femto Maximum Sensitivity Substrate, Thermo Fisher Scientific). Membrane blocking, secondary antibody incubation, washing and ECL incubation were performed at room temp, on a tipping board. Primary antibody incubation was performed at 4°C on a tipping board. Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA), and total protein content was quantified using ImageJ (RUEDENETAL2017), whereas total protein content was defined as mean grey value of the whole well with between-well values subtracted as background.

@@

At baseline, there were no significant differences between training with glucose or placebo in muscle mass, isometric or isokinetic peak torque, total session volume, training intensity, total RNA content, or rRNA expression (Tab 3). Further, on pairwise consecutive days, there were no significant differences in mean macro nutrient (calories, carbohydrate, fat or protein) ingestion (Tab 4), or mean change in macro nutrient ingestion (Fig 3). Resistance training with glucose led to significant increases in blood levels glucose and c-peptide compared to RT with placebo, before, during and after the session (Fig 4B, 120min: 38 ± 4% [*p* = `r p.120`], 135min: 31 ± 4% [*p* = `r p.135`], 150min: 32 ± 4% [*p* = `r p.150`], Fig 4C, 120min: 95 ± 10% [*p* = `r cpp.120`], 150min: 87 ± 10% [*p* = `r cpp.150`]), measured during the final RT session. At 2hrs post-exercise, RT with glucose led to significantly lower blood glucose levels compared to RT with placebo, with a tendancy of the same in c-peptide levels (4B, 270min: -8 ± 4% [*p* = `r p.270`] and 4C, 270min: -7 ± 10 [*p* `r cpp.270`]).

###### **Table 3**: Baseline characteristics of the intervention legs, organized per supplement. Variables: Iso 60 = isokinetic 60 d/s, Iso 240 = isokinetic 240 d/s, training intensity = load/rm*100. Nm = newton meters, kg = kilograms, %1RM = percentage of 1 repetition maximum, ng/mg = nanogram per milligram wet muscle tissue, nf.expr = expression normalized by weight and external reference gene (Lambda). Ribosomal RNA data are presented as log transformed. Glucose n = 13, placebo n = 13.

![](figures/table3.png)

###### **Table 4:** Daily mean dietary intake of macro nutrients during each training day per supplement leg. G = glucose, P = placebo. Training day 1-training day 12 are combined pairwise to day 1-6: Day 1 = training day 1/2, Day 2 = training day 3/4, Day 3 = training day 5/6, Day 4 = training day 7/8, Day 5 = training day 9/10, Day 6 = training day 11/12. Values are mean ± standard deviation. P = difference between glucose and placebo. Glucose n = 12, placebo n = 12. Day 6: Glu = 11, Plac = 11.

```{r nutrition table}
## Scripts
# - nut.table.R
# - nut.analysis.R
# - dxa.R

# Data
nuttable <- readRDS("./data/data-gen/dxa/nuttable.RDS")

nuttable %>%
  kable(col.names = c("Time", "Suppl.", "Kcal", "CHO (g)", "Fat (g)", "PRO (g)", "PRO (g/kg)"))

```{r nutrition change analysis and fig}
# Associated scripts:
# - nut.datahand.R
# - nut.change.R
# - nut.change.figs.R

nut.fig <- readRDS("./data/data-gen/nutrition/nut.fig.RDS")
nut.fig

```
###### **Figure 3:** Mean changes in macro nutrient intake from day to day. Day 1 = training day 1/2(T1/T2), Day 2 = training day 3/4, Day 3 = training day 5/6, Day 4 = training day 7/8, Day 5 = training day 9/10, Day 6 = training day 11/12(T3/T4). Values are mean change in grams/calories ± 95% CL.

![](figures/day.timeline.png)

```{r glucose + c-peptide fig}
## Scripts
# - gluc.fig.R
# - cpep.figs.R

# Separate figures
pos <- position_dodge(width = 0.2)

gluc.fig <- readRDS("./data/data-gen/glucose/gluc.fig.RDS")

cpep.fig <- readRDS("./data/data-gen/glucose/cpep.fig.RDS")

# Joined figure
plot_grid(gluc.fig,
          cpep.fig,
          ncol = 1, nrow = 2)

###### **Figure 4:** A) Time line for training day 6 (combination of pairwise training day 11 and 12). Biopsy = muscle biopsy taken pre supplementation, PRO = protein supplementation, Humac = strength test (isometric knee extension, isokinetic knee extension 60 and 240 d/s), GLU/PLAC = glucose/placebo supplementation, RT = resistance training. Minutes: -120 = 2hrs prior to RT, -90 = 90min prior to RT, -30 = 30min prior to RT, 0 = onset of RT, 15 = 15min following onset of RT, between leg press and knee extension, 30 = 30min post onset of RT, 120 = 2hrs following onset of RT. B) Mean mmol/L change in blood glucose levels per supplement group during training day 6. Time-points: 0, 45, 90, 120, 135, 150 and 270min post protein ingestion, respectively. C) Mean change in pmol/L of c-peptide in blood per supplement group during training day 6. Values are presented as estimated marginal means of mmol/L and pmol/L change ± 95% CI. *p* < 0.05 between groups. Glucose n = 13, placebo n = 13.

@@

Resistance training with glucose did not induce a higher accumulation of total RNA (fig 5A, *p* = `r trdiffp`), or rRNA  (fig 5B, *p* > 0.05), compared to RT with placebo, measured as mean change from baseline to post. From baseline to post-intervention, there was a mean increase in levels of total RNA by 26 and 22% after RT with glucose and placebo, respectively (*p* < 0.05, compared to baseline). A robust accumulation was also observed in ribosomal RNA expression in both RT with glucose and placebo, with mean increases between 34-43% (GLU) and 33-41% (PLAC) (*p* < 0.05, compared to baseline) in the four rRNA’s. The expression of 47S pre-rRNA increased by 37 and 59% with glucose and placebo respectively, with only placebo reaching significantly higher levels at post compared to baseline (fig 45, *p* = `r rr47tdiffp`).

>>>>

---
title: "results"
output: word_document
date: "2023-02-20"
editor_options: 
  chunk_output_type: console
---

@@

```{r packages}
library(tidyverse);library(readxl);library(lme4);library(emmeans);library(dplyr);library(cowplot);library(knitr);library(broom);library(rmarkdown);library(tinytex);library(tidyr);library(arsenal);library(magrittr);library(dabestr);library(nlme); library(pbkrtest);library(ggpubr)
```

@@

At baseline, there were no significant differences between training with glucose or placebo in muscle mass, isometric or isokinetic peak torque, total session volume, training intensity, total RNA content, or rRNA expression (Tab 3). Further, on pairwise consecutive days, there were no significant differences in mean macro nutrient (calories, carbohydrate, fat or protein) ingestion (Tab 4), or mean change in macro nutrient ingestion (Fig 3). Resistance training with glucose led to significant increases in blood levels glucose and c-peptide compared to RT with placebo, before, during and after the session (Fig 1B, 120min: 38 ± 4% [*p* = `r p.120`], 135min: 31 ± 4% [*p* = `r p.135`], 150min: 32 ± 4% [*p* = `r p.150`], Fig 1C, 120min: 95 ± 10% [*p* = `r cpp.120`], 150min: 87 ± 10% [*p* = `r cpp.150`]), measured during the final RT session. At 2hrs post-exercise, RT with glucose led to significantly lower blood glucose levels compared to RT with placebo (1B, 270min: -8 ± 4% [*p* = `r p.270`].

FIG1 HERE

**Figure 1:** A) Overview of the experimental design with 12 days of concomitant dietary intervention and resistance training (RT), preceded by 7 days of testing. Familiarization = days -7 to -1 included familiarization to RT exercises (via 1RM leg press and knee extension tests) and strength tests (Humac dynamometer) at days -7 and -5. Baseline strength tests were conducted at day -1. Biopsies were taken from vastus lateralis at baseline (Day 1 leg 1, Day 2 leg 2) and after five RT sessions (Day 11 leg 1, Day 12 leg 2). The order in which participants performed the two intervention blocks was determined in a planned randomized fashion, where one leg was randomized to RT with GLU (glucose) and the other to RT with PLA (placebo). The randomization also entailed half starting with GLU (n = 8) and the other half with PLA (n = 8), in addition they where further randomized to start either with their dominant or non-dominant leg. Thus, 4 participants were randomized to start RT with their dominant leg and GLU supplement, and 4 with their non-dominant leg and GLU - the same was true for participtants commencing the intervention with PLA. From intervention start (Day 1 leg 1, Day 2 leg 2), the participants exercised in a unilateral, alternating fashion until post intervention testing (Day 11/12 leg 1, Day 12/13 leg 2). Three participants dropped out of the intervention, either due to reported muscular discomforts to heavy strength training or sickness in the family. B and C) Changes in Plasma glucose (mmol/L) and -c-peptide levels (pmol/L) during RT session six (Day 11 leg 1, Day 12 leg 2). Finger draws and venous blood were sampled at baseline (Day -1), as well as before, during and after session six. Plasma glucose was measured via finger draws, 120 (-120), 90 (-90) and 30 (-30) minutes and immediately before RT (0), during RT (15), immediately after (30) and 2hrs after RT (120). Levels of c-peptide were measured via venous blood samples collected at the same time points as finger draws, except for 90min before RT and during RT. D, E, F) Changes in muscular performance (D: nm fold change in isometric peak torque, E/F: nm fold change in isokinetic peak torque at 60/240 d/s.). Muscular performance was measured via a Humac Norm Dynamometer at baseline (Day -1), during (Day 4 and 8 leg 1, Day 5 and 9 leg 2), and after five RT sessions (Day 11 leg 1, Day 12 leg 2). G, H, I) Changes in muscular recovery (G: nm fold change in isometric peak torque, H/I: nm fold change in isokinetic peak torque at 60/240 d/s). Muscular recovery was measured via Humac Norm Dynamometer with baseline being post fifth RT session test, and then 30min, 2hrs and 23hrs following the sixth and final RT session. Values are presented as estimated marginal means ± 95% CI. * = *p* < 0.05 between groups. Glucose n = 13, placebo n = 13.

@@

Resistance training with glucose did not induce a higher accumulation of total RNA (fig 2A, *p* = `r trdiffp`), or rRNA  (fig 2B-2F *p* > 0.05), compared to RT with placebo, measured as mean change from baseline to post. From baseline to post-intervention, there was a mean increase in levels of total RNA by 26 and 22% after RT with glucose and placebo, respectively (*p* < 0.05, compared to baseline). A robust accumulation was also observed in ribosomal RNA expression in both RT with glucose and placebo, with mean increases between 34-43% (GLU) and 33-41% (PLAC) (*p* < 0.05, compared to baseline) in the four rRNA’s. The expression of 47S pre-rRNA increased by 37 and 59% with glucose and placebo respectively, where only RT with placebo increased significantly from baseline to post (fig 45, *p* = `r rr47tdiffp`).

FIG2 HERE
